---
title: "Practical Machine learning Project"
author: "Cyprian Ngolah"
date: "January 22, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction 
One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise.


## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

## Data Loading and Processing
### Loading libraries

```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(RGtk2)
library(rattle)
library(randomForest)
library(repmis)
library(corrplot)
library(gbm)

training <- read.csv("pml-training.csv", na.strings = c("NA", ""))
testing <- read.csv("pml-testing.csv", na.strings = c("NA", ""))
```

The training dataset has 19622 observations and 160 variables, and the testing data set contains 20 observations and the same variables as the training set. We are trying to predict the outcome of the variable classe in the training set.

##Data cleaning
Columns with values from the training set will be deleted.
```{r}
training <- training[, colSums(is.na(training)) == 0]
testing <- testing[, colSums(is.na(testing)) == 0]
```
As the first 7 predictors have little predicting power, these are eleminated in the the training and testing sets.
```{r}
trainData <- training[, -c(1:7)]
testData <- testing[, -c(1:7)]
```
## Data spliting
In order to get out-of-sample errors, we split the cleaned training set trainData into a training set (train, 70%) for prediction and a validation set (valid 30%) to compute the out-of-sample errors.
```{r}
set.seed(7826) 
inTrain <- createDataPartition(trainData$classe, p = 0.7, list = FALSE)
train <- trainData[inTrain, ]
valid <- trainData[-inTrain, ]
```

##Prediction Algorithms

We will investigate the classification trees and random forests to predict the outcome.

###Classification trees
In practice, k=5 or k=10 when doing k-fold cross validation. Here we consider 5-fold cross validation (default setting in trainControl function is 10) when implementing the algorithm to save a little computing time. Since data transformations may be less important in non-linear models like classification trees, we do not transform any variables.

```{r}
control <- trainControl(method = "cv", number = 5)
fit_rpart <- train(classe ~ ., data = train, method = "rpart", 
                   trControl = control)
print(fit_rpart, digits = 4)

fancyRpartPlot(fit_rpart$finalModel)
```

```{r}
# predict outcomes using validation set
predict_rpart <- predict(fit_rpart, valid)
# Show prediction result
(conf_rpart <- confusionMatrix(valid$classe, predict_rpart))
```

From the confusion matrix, the accuracy rate is 0.5, and so the out-of-sample error rate is 0.5. Using classification tree does not predict the outcome classe very well.

###Random forests
Since classification tree method does not perform well, we try random forest method instead.
```{r}
fit_rf <- train(classe ~ ., data = train, method = "rf", 
                   trControl = control)
print(fit_rf, digits = 4)
```

```{r}
# predict outcomes using validation set
predict_rf <- predict(fit_rf, valid)
# Show prediction result
(conf_rf <- confusionMatrix(valid$classe, predict_rf))
```

```{r}
(accuracy_rf <- conf_rf$overall[1])
```

We can see from the results that for this dataset, random forest method is way better than classification tree method. The accuracy rate is 0.991, and so the out-of-sample error rate is 0.009. This may be due to the fact that many predictors are highly correlated. Random forests chooses a subset of predictors at each split and decorrelate the trees. This leads to high accuracy, although this algorithm is sometimes difficult to interpret and computationally inefficient.

###Prediction on Testing Set
With the above conlusion, We now use random forests to predict the outcome variable classe for the testing set.
```{r}
(predict(fit_rf, testData))
```

##Conclusion

Using the Random forest algorithm, the 20 predicted values are as shownb above.

